The effects of alignment error and alignment filtering on the \sw detection of positive selection

### 2011-07-02 projects/slrsim/paper/slrsim.tex
Paper draft.

\documentclass{mbe}
\usepackage{natbib}
\setlength{\bibsep}{0.0pt}

@@

\newcommand{\pranka}{PRANK\_A}
\newcommand{\prankc}{PRANK\_C}
\newcommand{\omg}{\bm{\omega}}
\newcommand{\tpr}{TPR$_{1\%}$}

\title{The effects of alignment error and alignment filtering on the sitewise detection of positive selection}

\author{{Gregory Jordan and Nick Goldman}\\
\affiliation{European Bioinformatics Institute, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}}
\authorlist{Jordan and Goldman}
\shorttitle{Alignment eror and detecting sitewise positive selection}

@@

The effects of alignment error on many types of evolutionary inference
are not well understood. Alignment filters are commonly applied to try
to reduce error rates resulting from misalignment, but neither the
quantitative impact of misalignment nor the ability of filters to
minimize such error has been systematically studied. We performed
simulation experiments to quantify the level of false positive and
false negative for sitewise detection of positive selection due to
misalignment and the ability of alignment filters to minimize
alignment-induced error. We find that while misalignment causes few
false positives when analyzing mammalian-like datasets with an
effective aligner, analyses at vertebrate-like or greater divergences
with either a poor aligner or a small number of sequences results in
high false positive rates.

@@

\keywords{Positive selection, multiple alignment, alignment filtering, likelihood.}

@@

The study of protein evolutionary rates and selective pressures in
particular has flourished as a result of the growth in comparative
genomics datasets. Clusters of closely-related genome sequences across
a wide taxonomic range have led to a better understanding of which
aspects of molecular evolution are variable and which are constant
\citep{Wolf2009Nonlinear}, and an increased sampling of species should
continue to boost the power and accuracy of individual analyses within
a given clade. This is especially beneficial for the calculation of
spatially precise evolutionary estimates, as additional species
sampling has been shown to be an effective means of boosting the
accuracy and power of sitewise detection of positive selection and
evolutionary constraint
\citep{Anisimova2001Accuracy,Massingham2005Detecting}. Site-specific
evolutionary estimates have proved especially valuable when analyzed
in conjunction with other protein-based datasets such as protein
structural features \citep{Lin2007Proportion,Ramsey2011Relationship},
human population diversity \citep{2010Map}, and human disease
mutations \citep{Arbiza2006Selective}.

A major concern in the sitewise detection of protein positive
selection is that the effect of alignment error is not well
characterized. Intuitively, one might expect alignment error to result
mainly in an increased number of false positives, as the spurious
alignment of nonhomologous codons on average would result in a high
number of apparent non-synonymous substitutions and a low number of
synonymous substitutions (since a randomly chosen pair of codons are
more likely to be nonsynonymous than synonymous). However, false
negatives may also be introduced, either through over-alignment of
synonymous codons (increasing the synonymous substitution rate) or
non-alignment of homologous codons (causing reduced power due to less
evolutionary information). Since different aligners employ a variety
of algorithms, evolutionary models, and heuristic optimizations
\citep{Notredame2007Recent}, each program may be more or less prone to different
types of alignment error, causing potentially large variations in the
nature and magnitude of its impact on the detection of positive
selection.

The protein structure and the evolutionary divergence of a dataset may
also contribute to the effects of alignment error. Differently
structured protein regions show variable tolerance to biological
indels, with indels more common in extracellular and transmembrane
proteins than in highly folded enzymes and housekeeping genes
\citep{delaChaux2007DNA} . This suggests that well-folded protein
regions will experience fewer biological indels -- and will therefore
be less susceptible to alignment error -- than less structured
regions. Furthermore, the evolutionary divergence of a dataset affects
the power of sitewise inference and the prevalence of indels in
different ways. Because maximum-likelihood methods for detecting
positive selection require data in the form of observed substitution
events, they show little power at low divergence and their highest
power at intermediate to high divergence levels
\citep{Anisimova2001Accuracy}. On the other hand, the number of indels
in a dataset should increase monotonically with divergence. These two
trends suggest that the overall power will be low at both extremes of
divergence, with little inference power at low divergences (due to the
scarcity of data in the form of observed substitutions) and an
overwhelming amount of alignment error at high divergences (due to the
large number of indel events). It is important to note, however, that
the majority of genes in many biological clades of interest (such as
mammals, vertebrates, fruit flies, and yeast) fall within the middle
range of divergences where sitewise methods are at their most powerful
and where multiple alignment is a difficult---but not
hopeless---problem. As such, it is important to seek an understanding
of the impact of alignment error on overall error rates within this
important range of divergence levels.

A number of empirical analyses have established that errors in gene
sequencing, annotation and alignment can contribute to errors in
downstream evolutionary analyses such as phylogeny inference
\citep{Wong2008Alignment} et and estimates of positive selection
\citep{Schneider2009Estimates,Markova-Raina2011High} . Most recently,
Markova-Raina et al. \citeyearpar{Markova-Raina2011High} showed that
the detection of positively-selected sites and genes in Drosophila
genomes is highly sensitive to aligner choice, with PRANK's codon
model consistently producing alignments with the lowest amount of
positive selection. Still, according to the authors' manual inspection
of alignments, even positively-selected sites identified with PRANK
alignments contained a sizable proportion of apparent false positives.

A limitation to the analysis of error in empirical datasets is the
lack of a benchmark set of true alignments and positively-selected
sites. Markova-Raina et al. \citeyearpar{Markova-Raina2011High} used
the expected general effect of alignment error (an increase in false
positives due to misalignment of non-homologous codons) as a proxy by
which to compare different methods, allowing for the conclusion that
PRANK was the least error-prone aligner in their analysis. However,
the absolute number of false positives remained uncertain due to the
possible conflation of multiple sources of error: in addition to
alignment error, the authors noted that gene mis-annotation was
responsible for many apparent false positives, and there was also an
expected error rate from the likelihood inference method itself. This
limitation leaves important and interesting questions regarding the
nature of misalignment error and its quantitative impact on the
detection of positive selection left unanswered.

Controlled simulation experiments provide a natural framework for
investigating error rates in detail, allowing one to pinpoint the
source of errors in multi-step analyses such as alignment followed by
evolutionary inference. This approach has widely been employed in
assessing the robustness of phylogenetic inference methods to
misalignment
\citep{Dwivedi2009Phylogenetic,Ogden2006Multiple,Loytynoja2008PhylogenyAware}
but those results cannot be easily extrapolated to the analysis of
sitewise selective pressures. More recently, Fletcher and Yang
\citeyearpar{Fletcher2010Effect} performed a series of simulation
experiments investigating alignment error in the use of branch-site
models to detect positive selection in genes. Their results showed
that most aligners caused false positives by over-aligning codons, and
that datasets from mammalian and vertebrate gene families contain
enough evolutionary divergence to make errors resulting from
misalignment a legitimate concern.

Reflecting a widespread awareness of the problem of misalignment,
methods for identifying and removing uncertain or unreliable alignment
regions have been commonly used in phylogenetic and molecular
evolutionary analyses. The popular Gblocks applies a set of heuristic
criteria (including thresholds on the amount of conservation and
gappiness) to identify conserved blocks deemed suitable for
phylogenetic or evolutionary analysis \citep{Castresana2000Selection},
while aligners such as T-Coffee \citep{Notredame2000TCoffee} and PRANK
\citep{Loytynoja2005From} produce estimates of alignment confidence or
reliability. Most recently, alignment confidence scores based on the
robustness of alignments to guide tree perturbations
\citep{Penn2010Alignment} or the sampling of suboptimal alignments
\citep{Kim2011PSAR} have been proposed. Unfortunately, despite their
widespread use, the impact of the many available filtering methods on
phylogenetic and evolutionary analyses has not been well-studied. Even
for a single filtering program, Gblocks, results have been
contradictory: one simulation-based study found that it improved the
phylogenetic signal \citep{Talavera2007Improvement} while an empirical
study across a wide range of taxa found that Gblocks-filtered
alignments produced worse phylogenetic trees
\citep{Dessimoz2010Phylogenetic}. The effect of filtering alignments
on the detection of positive selection is largely unexplored, although
it has become accepted practice to apply published filtering methods
before testing for positive selection \citep{Studer2008Pervasive,Aguileta2009Rapidly}.

This paper aims to use a simulation framework to incorporate
misalignment and alignment filtering into estimates of the power and
accuracy of sitewise evolutionary estimates. Our approach is similar
to that of Anisimova and Yang \citeyearpar{Anisimova2002Accuracy} in
that we use simulated protein alignments to evaluate methods for
detecting sitewise positive selection, considering each site to be an
independent hypothesis test producing a positive or negative
result. Our experimental design also shares some features with
Fletcher and Yang's \citeyearpar{Fletcher2010Effect} investigation of
the performance of branch-site methods for detecting genewise positive
selection under alignment error. Like Fletcher and Yang
\citeyearpar{Fletcher2010Effect}, we simulate codon alignments with
indels and calculate inferred alignments using a small but diverse
sample of aligners. Unlike their study, however, we focus on sitewise
detection of positive selection occurring throughout a phylogeny (as
opposed to genewise detection of selection acting at specific
branches), we explore a wider range of plausible tree sizes, indel
rates and divergence levels, and we evaluate the effects of a number
of alignment filtering methods on the error rate and power of sitewise
analysis. Given the recent proliferation of completed and planned
vertebrate genomes, we pay special attention to choosing simulation
parameters and analysis methods similar to those commonly encountered
in the sitewise analysis of vertebrate gene families.

@@

@@

Three rooted trees were used to guide the simulation of protein-coding
DNA alignments: the artificial 6-taxon tree used by Anisimova and Yang
\citeyearpar{Anisimova2001Accuracy} and Massingham and Goldman
\citeyearpar{Massingham2005Detecting} rooted at its midpoint, the
17-taxon vertebrate B-globin tree from Yang et
al. \citeyearpar{Yang2000CodonSubstitution}, and the 44-taxon
vertebrate tree used by the ENCODE project
\citep{2007Identification,Nikolaev2007Early}. Trees, shown with their
original branch lengths in Figure 1, were scaled to comparable
divergence levels by normalizing their root-to-tip mean path length
(MPL), defined as the root-to-tip branch length averaged across all
lineages in the tree. We simulated alignments with MPL divergence
between 0.2 and 2.0 synonymous substitutions per synonymous site,
spanning the range of evolutionary divergences observed in several
clades of organisms with fully-sequenced genomes (Table ST1).

\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.5]{fig1.pdf}
\end{center}
\caption{Phylogenetic trees used for simulation and analysis in this
  article. (a) An artificial tree used in previous simulations
  \citep{Anisimova2001Accuracy,Massingham2005Detecting}. (b) A tree
  estimated from b-globin genes of 17 vertebrates and used in previous
  empirical analyses and simulation studies
  \citep{Anisimova2001Accuracy,Anisimova2002Accuracy}. (c) The
  44-taxon tree used by the ENCODE project
  \citep{2007Identification,Nikolaev2007Early}. The nodes labeled in
  red were used as the reference species (see Methods).}
\label{fig_1}
\end{figure}

The Indelible program \citep{Fletcher2009INDELible} was used to
simulate codon sequences with indels along each phylogenetic tree. The
length of the root sequence was set to 500 codons, $\bm{\kappa}$ (the
ratio of transition to transversion substitutions) was fixed at 4, and
100 replicate alignments were simulated for each set of
conditions. Indel lengths were drawn from a discretized power-law
distribution with an exponential decay parameter of 1.8 and a maximum
value of 40, yielding a mean indel length of 3.33 codons and standard
deviation of 5.51 codons. This power-law model of indel lengths is
well-supported by empirical studies
\citep{Benner1993Empirical,Cartwright2009Problems} and it was decided
after manual inspection of alignments from a range of parameter values
that the chosen model parameters resulted in alignments most closely
resembling those encountered in vertebrate alignments. The ratio of
insertion to deletion events was set to 1, and the rate of indel
formation was scaled from between 0 to 0.2 indel events per
substitution per site.

The distribution of sitewise selective pressures was modeled with a
log-normal distribution derived from a maximum-likelihood fit to a
large dataset of sitewise selective pressures estimated from mammalian
gene trees (log-normal parameters shown in Table 1). This
distribution, with mean w of 0.28 and 6\% of sites having w>1, is
consistent with the structure-based expectation of many protein sites
under purifying selection and few under neutral selection or positive
selection
\citep{Smith1970Natural,Kimura1974SomePrinciples}. Indelible's general
discrete model of sitewise w variation was used to approximate this
distribution by splitting the lognormal probability density into 50
equally-spaced bins between w values of 0 and 3, with the highest bin
containing the probability density for all values w>=3.

Branch lengths for each of the simulation trees were multiplied by a
correction factor before simulation to correct for the difference
between our definition of branch lengths as the number of synonymous
substitutions per synonymous site ($dS$) and Indelible's
interpretation of branch length as the average number of substitutions
per codon ($t$) \citep{Fletcher2010Effect}. They are related
approximately by $t = 3(NdN + SdS) = 3dS(\omg{}N + S)$, where $N$ and
$S$ are the proportion of nonsynonymous and synonymous sites and
$\omg{}$ is the mean $\omg{}$ across all sites. S is approximately 0.3
when $\bm{\kappa}=4$ and the mean $\omg{}$ ratio for our chosen
distribution is 0.277, yielding a $dS$-to-$t$ conversion factor of
1.48 for all simulations performed.

@@

Alignments were filtered by masking out residues based on the output
of four alignment scoring methods: Gblocks conserved blocks
\citep{Castresana2000Selection}, T-Coffee consistency scores
\citep{Notredame2000TCoffee,Notedame2003Using}, and GUIDANCE alignment
confidence scores \citep{Penn2010Alignment}. Gblocks, which identifies
entire alignment columns as conserved or not conserved, was run using
a relaxed gap allowance parameter in order to avoid removing too much
of the alignment (command-line parameter 'b5=a') and all residues from
any columns not within an identified conserved block were masked with
'N's. GUIDANCE and T-Coffee, which produce individual scores for each
residue, were handled differently. (TODO!!!!). GUIDANCE was run with 30
bootstrap replicates and MAFFT as the bootstrap aligner (except when
\pranka{} or \prankc{} was the original aligner, in which case \pranka{}
was used to align the bootstrap replicates), and T-Coffee was run
using default settings and the 'evaluate\_mode -output=score\_ascii'
command-line parameter to output alignment scores.

Two unrealistic datasets were produced to serve as controls. First,
the true simulated alignment was used in order to evaluate the
sitewise performance without any alignment error. Second, an
additional filtering method was designed to represent an unattainable
best-case scenario for sequence filtering, using knowledge of the true
alignment to assign a score to each residue that summarizes how
correctly is has been placed in the inferred alignment. The approach
taken was to calculate, for each residue, the branch length of the
correct sub-tree (defined as the minimum sub-tree connecting all
sequences to which the current residue is correctly aligned) divided
by the branch length of the aligned sub-tree (defined as the minimum
sub-tree connecting all sequences with non-gap residues at the current
alignment column). This residue-wise score ranges from 0 to 1 and
encapsulates the expectation that correctly-aligned evolutionary
branch length is the main source of information from which the
sitewise inference methods derive their power. Scores from this
method, referred to as the Optimal method, were used to filter
inferred alignments in a manner similar to the T-Coffee and GUIDANCE
scores.

\subsection*{Sitewise Evolutionary Analysis}

Sitewise estimates of selective pressures were calculated using
maximum-likelihood methods implemented in the Phylogenetic Analysis by
Maximum Likelihood (PAML, \citealt{Yang2007PAML}) and Sitewise Likelihood Ratio
(SLR, \citealt{Massingham2005Detecting}) software packages.

The codeml program from PAML implements a number of likelihood ratio
tests (LRTs) for detecting the presence of positive selection in a
gene while allowing the $\omg{}$ ratio to vary among sites
\citep{Yang2000CodonSubstitution}. These models, known as the sites or
random sites models, use a variety of predefined statistical
distributions to account for heterogeneous w ratios among sites. After
the likelihood optimization is performed, Bayesian methods can be used
to estimate the posterior probability of each site being drawn from a
given site class, where a high posterior probability of a site
belonging to a class with $\omg{}>1$ can be considered strong
evidence that a site has evolved under positive selection
\citep{Yang2005Bayes}. We used the two models for which the
recommended Bayes Empirical Bayes method are implemented, M2a and M8.

SLR implements a method specifically designed for sitewise estimates
which has been shown in simulations to perform as well or better than
PAML's sitewise models \citep{Massingham2005Detecting}. SLR models
codon evolution as a continuous-time Markov process where
substitutions at one site are independent of substitutions at all
other sites. Importantly, no assumptions are made regarding the
distribution of $\omg{}$ ratios within the alignment. The value
of $\omg{}$ is considered to be an independent parameter at each
site: after first optimizing shared parameters using the whole
alignment, SLR uses the shared parameters and the data at each
alignment site to calculate a sitewise statistic for non-neutral
evolution. This statistic is based on a likelihood-ratio test where
the null model is neutral evolution ($\omg{}=1$) and the
alternative model is either purifying or positive selection
($\omg{}<1$ or $\omg{}>1$, respectively). The raw statistic
measures the strength of evidence for non-neutral evolution at each
site; following Massingham \citeyearpar{Massingham2005Detecting} we
use a signed version of the SLR statistic (created by negating the
statistic for sites with $\omg{}<1$) as the test statistic for
positive selection.

>>>>

\documentclass{mbe}
\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
\usepackage{xspace}
\setlength{\bibsep}{0.0pt}

@@

\newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}
\newcommand{\prankc}{PRANK$_{\textrm{C}}$\xspace}
\newcommand{\omg}{\bm{\omega}\xspace}
\newcommand{\tpr}{TPR$_{1\%}$\xspace}
\newcommand{\sw}{sitewise\xspace}
\newcommand{\Sw}{Sitewise\xspace}
\newcommand{\Dr}{{\em Drosophila}\xspace}

\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection}

\author{{Gregory Jordan and Nick Goldman}\\
\affiliation{European Bioinformatics Institute, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}}
\authorlist{Jordan and Goldman}
\shorttitle{Effect of Alignment Error on Detecting Sitewise Selection}

@@

The effects of alignment error on many types of evolutionary inference
are not well understood. Alignment filters are commonly applied to try
to reduce errors resulting from misalignment, but neither the
prevalence of such errors nor the ability of filters to reduce them
has been systematically studied. Focusing on the sitewise detection of
positive selection, we performed simulation experiments to quantify
the false positives and false negatives introduced by alignment error
and the ability of alignment filters to improve performance. We found
that some aligners produced many false positive results across a wide
range of divergence levels, while others produced very few. False
negatives were a problem for all aligners, increasing with
divergence. Of the aligners tested, PRANK consistently performed the
best and ClustalW performed the worst. The best alignment filters
showed good ability to reduce the error rates from ClustalW
alignments, but they failed to improve the performance of PRANK
alignments under most circumstances. Of the filters tested, GUIDANCE
performed the best and Gblocks performed the worst, failing to improve
even the error-prone ClustalW alignments. The performance of the true
alignment and an optimal filtering method suggested that performance
improvements could be gained by improving aligners or filters to
reduce the prevalence of false negatives, especially at higher
divergence levels and indel rates. We conclude that for the sitewise
evolutionary analysis of most vertebrate, insect and fungal
phylogenies, the use of unfiltered PRANK alignments is recommended.

@@

\keywords{Alignment, positive selection, alignment filtering, indels}

@@

The study of protein evolutionary rates and selective pressures in
particular has flourished as a result of the growth in comparative
genomics datasets. Clusters of closely-related genome sequences across
a wide taxonomic range have led to a better understanding of which
aspects of molecular evolution are variable and which are constant
\citep{Wolf2009Nonlinear}, and an increased sampling of species should
continue to boost the power and accuracy of individual analyses within
a given clade. This is especially beneficial for the calculation of
spatially precise evolutionary estimates, as additional species
sampling has been shown to be an effective means of boosting the
accuracy and power of \sw detection of positive selection and
evolutionary constraint
\citep{Anisimova2001Accuracy,Massingham2005Detecting}. Site-specific
evolutionary estimates have proved especially valuable when analyzed
in conjunction with other protein-based datasets such as protein
structural features \citep{Lin2007Proportion,Ramsey2011Relationship},
human population diversity \citep{2010Map}, and human disease
mutations \citep{Arbiza2006Selective}.

A major concern in the \sw detection of protein positive
selection is that the effect of alignment error is not well
characterized. Intuitively, one might expect alignment error to result
mainly in an increased number of false positives, as the spurious
alignment of nonhomologous codons on average would result in a high
number of apparent non-synonymous substitutions and a low number of
synonymous substitutions (since a randomly chosen pair of codons are
more likely to be nonsynonymous than synonymous). However, false
negatives may also be introduced, either through over-alignment of
synonymous codons (increasing the synonymous substitution rate) or
non-alignment of homologous codons (causing reduced power due to less
evolutionary information). Since different aligners employ a variety
of algorithms, evolutionary models, and heuristic optimizations
\citep{Notredame2007Recent}, each program may be more or less prone to different
types of alignment error, causing potentially large variations in the
nature and magnitude of its impact on the detection of positive
selection.

\input{table1.txt}

The protein structure and the evolutionary divergence of a dataset may
also contribute to the effects of alignment error. Differently
structured protein regions show variable tolerance to biological
indels, with indels more common in extracellular and transmembrane
proteins than in highly folded enzymes and housekeeping genes
\citep{delaChaux2007DNA} . This suggests that well-folded protein
regions will experience fewer biological indels---and will therefore
be less susceptible to alignment error---than less structured
regions. Furthermore, the evolutionary divergence of a dataset affects
the power of \sw inference and the prevalence of indels in different
ways. Because maximum-likelihood methods for detecting positive
selection require data in the form of observed substitution events,
they show little power at low divergence and their highest power at
intermediate to high divergence levels
\citep{Anisimova2001Accuracy}. On the other hand, the number of indels
in a dataset should increase with divergence. These two trends suggest
that the overall power will be low at both extremes of divergence,
with little inference power at low divergence (due to the scarcity of
data in the form of observed substitutions) and an overwhelming amount
of alignment error at high divergence (due to the large number of
indel events). It is important to note, however, that the majority of
genes in many biological clades of interest (such as mammals,
vertebrates, fruit flies, and yeast) fall within the middle range of
divergences where \sw methods are at their most powerful and where
multiple alignment is a difficult---but not hopeless---problem. As
such, it is important to seek an understanding of the impact of
alignment error on overall error rates within this important range of
divergence levels.

A number of empirical analyses have established that errors in gene
sequencing, annotation and alignment can contribute to errors in
downstream evolutionary analyses such as phylogeny inference
\citep{Wong2008Alignment} et and estimates of positive selection
\citep{Schneider2009Estimates,Markova-Raina2011High} . Most recently,
Markova-Raina and Petrov \citeyearpar{Markova-Raina2011High} showed
that the detection of positively-selected sites and genes in \Dr
genomes is highly sensitive to aligner choice, with PRANK's codon
model consistently producing alignments with the lowest amount of
positive selection. Still, according to the authors' manual inspection
of alignments, even positively-selected sites identified with PRANK
alignments contained a sizable proportion of apparent false positives.

A limitation to the analysis of error in empirical datasets is the
lack of a benchmark set of true alignments and positively-selected
sites. Markova-Raina and Petrov \citeyearpar{Markova-Raina2011High}
used the expected general effect of alignment error (an increase in
false positives due to misalignment of non-homologous codons) as a
proxy by which to compare different methods, allowing for the
conclusion that PRANK was the least error-prone aligner in their
analysis. However, the absolute number of false positives remained
uncertain due to the possible conflation of multiple sources of error:
in addition to alignment error, the authors noted that gene
mis-annotation was responsible for many apparent false positives, and
there is also an expected error rate from the likelihood inference
method itself. This limitation leaves important and interesting
questions regarding the nature of alignment error and its quantitative
impact on the detection of positive selection left unanswered.

Controlled simulation experiments provide a natural framework for
investigating error rates in detail, allowing one to pinpoint the
source of errors in multi-step analyses such as alignment followed by
evolutionary inference. This approach has widely been employed in
assessing the robustness of phylogenetic inference methods to
misalignment
\citep{Dwivedi2009Phylogenetic,Ogden2006Multiple,Loytynoja2008PhylogenyAware}
but those results cannot be easily extrapolated to the analysis of \sw
selective pressures. More recently, Fletcher and Yang
\citeyearpar{Fletcher2010Effect} performed a series of simulation
experiments investigating alignment error in the use of branch-site
models to detect positive selection in genes. Their results showed
that most aligners caused false positives by over-aligning codons and
that datasets from mammalian and vertebrate gene families contain
enough evolutionary divergence to make false positive errors resulting
from misalignment a legitimate concern.

Reflecting a widespread awareness of the problem of misalignment,
methods for identifying and removing uncertain or unreliable alignment
regions have been commonly used in phylogenetic and molecular
evolutionary analyses. The popular Gblocks applies a set of heuristic
criteria to identify conserved blocks deemed suitable for phylogenetic
or evolutionary analysis \citep{Castresana2000Selection} while a
number of aligners such as T-Coffee \citep{Notredame2000TCoffee} and
PRANK \citep{Loytynoja2005From} produce estimates of alignment
confidence or reliability. Alignment confidence scores based on the
robustness of alignments to guide tree perturbations
\citep{Penn2010Alignment} or the sampling of suboptimal alignments
\citep{Kim2011PSAR} have also been proposed. Unfortunately, despite
their widespread use, the impact of the many available filtering
methods on phylogenetic and evolutionary analyses has not been well
studied. Even for a single filtering program, Gblocks, results have
been contradictory: one simulation-based study found that it improved
the phylogenetic signal \citep{Talavera2007Improvement} while an
empirical study across a wide range of taxa found that
Gblocks-filtered alignments produced worse phylogenetic trees than
unfiltered alignments \citep{Dessimoz2010Phylogenetic}. With the
application of published filtering methods to alignments before
testing for positive selection becoming standard practice
\citep{Studer2008Pervasive,Aguileta2009Rapidly}, an analysis of the
benefit---if any---of alignment filtering to the detection of positive
selection would seem well-warranted.

This paper aims to use a simulation framework to incorporate alignment
error and alignment filtering into estimates of the error rate and
power of \sw evolutionary estimates. Our approach is similar to that
of Anisimova and Yang \citeyearpar{Anisimova2002Accuracy} in that we
use simulated protein alignments to evaluate methods for detecting \sw
positive selection, considering each site to be an independent
hypothesis test producing a positive or negative result. Our
experimental design also shares some features with Fletcher and Yang's
\citeyearpar{Fletcher2010Effect} investigation of the performance of
branch-site method for detecting genewise positive selection under
alignment error. Like Fletcher and Yang
\citeyearpar{Fletcher2010Effect}, we simulate codon alignments with
indels and calculate inferred alignments using a small but diverse
sample of aligners. Unlike their study, here we focus on \sw detection
of positive selection occurring throughout a phylogeny (as opposed to
genewise detection of selection acting at specific branches), we
explore a wider range of plausible tree sizes, indel rates and
divergence levels, and we evaluate the impact of a number of alignment
filtering methods on the \sw analysis. Given the recent proliferation
of completed and planned vertebrate genomes, we pay special attention
to choosing simulation parameters and analysis methods similar to
those commonly encountered in the \sw analysis of vertebrate gene
families.

@@

\input{table2.txt}

@@

An overview of the simulation parameters used in this study can be
found in Table \ref{table_1}. Three rooted trees were used to guide
the simulation of protein-coding DNA alignments: the artificial
6-taxon tree used by Anisimova and Yang
\citeyearpar{Anisimova2001Accuracy} and Massingham and Goldman
\citeyearpar{Massingham2005Detecting} rooted at its midpoint, the
17-taxon vertebrate B-globin tree from Yang et
al. \citeyearpar{Yang2000CodonSubstitution}, and the 44-taxon
vertebrate tree used by the ENCODE project
\citep{2007Identification,Nikolaev2007Early}. Trees, shown with their
original branch lengths in Figure \ref{fig_1}, were scaled to
comparable divergence levels by normalizing their mean path length
(MPL), defined as the root-to-tip branch length averaged across all
lineages in the tree. We simulated alignments with MPL divergence
between 0.05 and 2.0 synonymous substitutions per synonymous site,
spanning the range of evolutionary divergences observed in several
clades of organisms with fully-sequenced genomes (Table
\ref{table_2}).
\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.5]{fig1.pdf}
\end{center}
\caption{Phylogenetic trees used for simulation and analysis. The
  original scale for each tree is indicated by a scale bar, but trees
  were scaled to equal mean path length (MPL) divergence levels for
  simulation. (a) A 6-taxon artificial tree used in previous
  simulations
  \citep{Anisimova2001Accuracy,Massingham2005Detecting}. (b) A tree
  estimated from $\beta$-globin genes of 17 vertebrates and used in
  previous empirical analyses and simulation studies
  \citep{Anisimova2001Accuracy,Anisimova2002Accuracy}. (c) The
  44-species tree used by the ENCODE project
  \citep{2007Identification,Nikolaev2007Early}. The nodes indicated by
  arrows were used as the reference species when comparing the true
  and inferred alignment (see Methods).}
\label{fig_1}
\end{figure}

The Indelible program \citep{Fletcher2009INDELible} was used to
simulate codon sequences with indels along each phylogenetic tree. The
length of the root sequence was set to 500 codons, $\bm{\kappa}$ (the
ratio of transition to transversion substitutions) was fixed at 4, and
100 replicate alignments were simulated for each set of conditions
unless otherwise noted. Indel lengths were drawn from a discretized
power-law distribution with an exponential decay parameter of 1.8 and
a maximum value of 40, yielding a mean indel length of 3.33 codons and
standard deviation of 5.51 codons. This power-law model of indel
lengths is well-supported by empirical studies
\citep{Benner1993Empirical,Cartwright2009Problems} and manual
inspection of alignments from a range of parameter values identified
the chosen model parameters as resulting in alignments most closely
resembling those encountered in vertebrate alignments. The ratio of
insertion to deletion events was set to 1, and the rate of indel
formation was scaled between 0 to 0.2 indel events per substitution
per site.

The distribution of \sw selective pressures (embodied by the parameter
$\omg$, the ratio of the rate of non-synonymous substitution to the
rate of synonymous substitution) was modeled with a log-normal
distribution derived from a maximum-likelihood fit to a large dataset
of \sw selective pressures estimated from mammalian gene trees
(unpublished data; log-normal parameters shown in Table
\ref{table_1}). This distribution, with mean $\omg$ of 0.28 and 6\% of
sites having $\omg>1$, is consistent with the structure-based
expectation of many protein sites under purifying selection and few
under neutral selection or positive selection
\citep{Smith1970Natural,Kimura1974SomePrinciples}. Indelible's general
discrete model of \sw $\omg$ variation was used to approximate the
lognormal distribution by splitting the probability density into 50
equally-spaced bins between $\omg$ values of 0 and 3, with the highest
bin containing the probability density for all values $\omg>3$.

Branch lengths for each of the simulation trees were multiplied by a
correction factor before simulation to correct for the difference
between our definition of branch lengths as the number of synonymous
substitutions per synonymous site ($dS$) and Indelible's
interpretation of branch length as the average number of substitutions
per codon ($t$) \citep{Fletcher2010Effect}. They are related
approximately by $t=3(NdN+SdS) = 3dS(\bar{\omg{}}N+S)$, where $N$ and
$S$ are the proportion of nonsynonymous and synonymous sites and
$\bar{\omg}$ is the mean $\omg$ across all sites. $S$ is approximately 0.3
when $\bm{\kappa}=4$ \citep{Yang1998Synonymous} and the mean $\omg$ ratio for our chosen
distribution is 0.277, yielding a $dS$-to-$t$ conversion factor of
1.48 for all simulations performed.

@@

Alignments were filtered by masking out residues based on the output
of three alignment scoring methods: Gblocks conserved blocks
\citep{Castresana2000Selection}, T-Coffee consistency scores
\citep{Notredame2000TCoffee,Notredame2003Using}, and GUIDANCE
alignment confidence scores \citep{Penn2010Alignment}. Gblocks, which
identifies entire alignment columns as conserved or not conserved, was
run using an increased gap tolerance and a reduced minimum block
length in order to reduce the amount of each alignment removed
(command-line parameters {\em b5=a} and {\em b4=3}), and all residues
from any columns not within an identified conserved block were masked
with $N$s. GUIDANCE and T-Coffee, which produce individual scores for
each residue, were handled differently. A cutoff threshold was chosen
for each method (0.5 for GUIDANCE and 5 for T-Coffee) and residues
equal to or below that threshold were masked. If the cutoff threshold
caused greater than 50\% of an alignment's residues to be masked, the
threshold was relaxed to the highest value for which at least 50\% of
residues remained unmasked. This cap on the severity of each filter
was necessary because the scores from both methods were strongly
affected by the simulation conditions, with much lower average scores
at higher indel rates and divergences. GUIDANCE was run with 100
bootstrap replicates and MAFFT as the bootstrap aligner (except when
\pranka{} or \prankc{} was the original aligner, in which case 30
bootstrap replicates were generated with \pranka{}) and T-Coffee was
run using default settings and the {\em evaluate\_mode
  -output=score\_ascii} command-line parameters to output alignment
scores.

Two unrealistic datasets were produced to serve as controls. First,
the true simulated alignment was included in order to evaluate the \sw
performance without any alignment error. Second, an additional
filtering method was constructed to represent an unattainable
best-case scenario for sequence filtering, using knowledge of the true
alignment to assign a score to each residue reflecting how correctly
is has been placed in the inferred alignment. The approach taken was
to calculate, for each residue, the branch length of the correct
sub-tree (defined as the sub-tree connecting all sequences to which
the current residue was correctly aligned) divided by the branch
length of the total aligned sub-tree (defined as the sub-tree
connecting all sequences with non-gap residues at the current
alignment column). This residue-wise score ranges from 0 to 1 and
reflects the expectation that correctly-aligned evolutionary branch
length is the main source of information from which \sw inference
methods derive their power. We refer to this method as the optimal
method. Scores were handled in a manner similar to GUIDANCE and
T-Coffee, using a score threshold of 0.5.

\subsection*{\Sw Evolutionary Analysis}

\Sw estimates of selective pressures were calculated using
maximum-likelihood methods implemented in the Phylogenetic Analysis by
Maximum Likelihood (PAML, \citealt{Yang2007PAML}) and Sitewise Likelihood Ratio
(SLR, \citealt{Massingham2005Detecting}) software packages.

The $codeml$ program from PAML implements a number of likelihood ratio
tests (LRTs) for detecting the presence of positive selection in a
gene while allowing the $\omg$ ratio to vary among sites
\citep{Yang2000CodonSubstitution}. These models, known as the sites or
random sites models, use a variety of predefined statistical
distributions to account for heterogeneous $\omg$ ratios among sites. After
the likelihood optimization is performed, Bayesian methods can be used
to estimate the posterior probability of each site being drawn from a
given site class, where a high posterior probability of a site
belonging to a class with $\omg>1$ can be considered strong
evidence that a site has evolved under positive selection
\citep{Yang2005Bayes}. We used the two models for which the
recommended Bayes Empirical Bayes method are implemented, M2a and M8.
